---
output:
  html_document:
    toc: true
    toc_float: true
    highlight: zenburn
    theme: flatly
---

```{r read-data}
library(readr)
bc = read_delim("r1.barcodes", delim=" ")
```

```{r barcode-group}
library(dplyr)
grouped = bc %>% group_by(barcode, index) %>% summarise(count=n())
```

# Summary stats
There are `r nrow(grouped)` unique barcode-index pairs and of those
`r nrow(subset(grouped, count > 10))` are seen more than 10 times.

We can see below that there is are a small set of barcode-index
pairs that are represented a high number of times.

```{r barcode-index-plot}
library(ggplot2)
ggplot(grouped, aes(count)) + geom_histogram() + theme_bw() +
  scale_x_log10() + scale_y_sqrt()
```

# Known barcodes
Now we'll mark up the barcode-index pair with whether or not these
are some of the known sequences. The indexes we found correspond to the
reverse complement column in the index file and the barcodes correspond
to the sense column in the barcode file.

```{r read-known-data}
kbc = read_csv("metadata/BC sequences.csv")
kin = read_csv("metadata/indexseqs.csv")
grouped$kin = grouped$index %in% kin$rc
grouped$kbc = grouped$barcode %in% kbc$sense
grouped$known = grouped$kin & grouped$kbc
grouped$known = ifelse(grouped$known, "known", "unknown")
known = subset(grouped, known == "known")
```

Now we can see if we mark up the barcode-index pairs with whether or not
they are from the known barcode and known index files, we find that
there are only `r nrow(known)` barcode-index pairs.

If we plot the histograms separating out whether they are a known or unknown
barcode-index pair, we can see that the known barcode-index pairs
have many more reads associated with them than the unknown barcode-index
pairs.

```{r known-plot}
library(ggplot2)
ggplot(grouped, aes(count)) + geom_histogram() + theme_bw() +
  facet_wrap(~known) +
  scale_x_log10() + scale_y_sqrt() +
  ylab("number of barcode-index pairs")
```

# Barcodes per index
We're expecting 2 to 3 barcodes per index, but we generally see more than that
if we're seeing `r nrow(known)` barcodes. How many per barcode?

```{r barcodes-per-index-plot}
barcode_per_index = known %>% group_by(index) %>% summarise(nbarcodes = n())
index_per_barcode = known %>% group_by(barcode) %>% summarise(nindex = n())
ggplot(barcode_per_index, aes(index, nbarcodes)) +
  geom_bar(stat='identity') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("")
ggplot(index_per_barcode, aes(barcode, nindex)) + geom_bar(stat='identity') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("")
```

# Overrepresented barcodes

We see every index with every barcode at least once. Are some overrepresented?

```{r overrpresented}
ggplot(known, aes(barcode, count)) + facet_wrap(~index) + geom_bar(stat='identity') +
  scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

For some indexes it looks like yes, and for others no.

# Filtering for barcode pairs
Here I filtered by keeping only the barcodes for each index which were at
least twice as prevalent as the average of the barcodes for that index. Then
I filtered out all indexes that had more than 3 barcodes that passed that
filter.

```{r filtering}
filtered = known %>% group_by(index) %>%
  mutate(mcount = mean(count)) %>%
  filter(count > 2*mcount) %>% group_by(index) %>%
  mutate(nfiltered = n()) %>%
  filter(nfiltered < 4)
```

That leaves us with these:

```{r filter-plot}
ggplot(filtered, aes(barcode, count)) + facet_wrap(~index) +
  geom_bar(stat='identity') + scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```


# Expected indexes
```{r expected-indexes}
expected = read_csv("metadata/expected-pairs.csv")
colnames(expected) = c("ebarcode", "eindex")
expected$pair = paste(expected$ebarcode, expected$eindex, sep=":")
filtered$pair = paste(filtered$barcode, filtered$index, sep=":")
filtered$expected = filtered$pair %in% expected$pair
```

```{r filter-plot-expected}
ggplot(filtered, aes(barcode, count, fill=expected)) + facet_wrap(~index) +
  geom_bar(stat='identity') + scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
output = filtered %>% select(barcode, index, count)
write.table(output, file="barcode-index.csv", col.names=TRUE,
            sep=",", quote=FALSE, row.names=FALSE)
```

# Expected on unfiltered list

```{r known-plot-expected}
known$pair = paste(known$barcode, known$index, sep=":")
known$expected = known$pair %in% expected$pair
ggplot(known, aes(barcode, count, fill=expected)) +
  facet_wrap(~index) + geom_bar(stat='identity') +
  scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

# 33 barcode mixes
```{r 33-barcode-alone}
index_mixes = c("ATAACGGT", "TCAGCATT", "TGTGACTA")
ggplot(subset(known, index %in% index_mixes),
       aes(barcode, count, fill=expected)) +
  facet_wrap(~index) + geom_bar(stat='identity') +
  scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

# Output tables
```{r output-tables}
output = filtered %>% select(barcode, index, count, expected)
write.table(output, file="filtered-barcode-index.csv", col.names=TRUE,
            sep=",", quote=FALSE, row.names=FALSE)
output = known %>% select(barcode, index, count, expected)
write.table(output, file="unfiltered-barcode-index.csv", col.names=TRUE,
            sep=",", quote=FALSE, row.names=FALSE)
```
[filtered-barcode-index.csv](filtered-barcode-index.csv)

[unfiltered-barcode-index.csv](unfiltered-barcode-index.csv)

I included an example of a CSV file of the expected pairs that is much easier
to work with:

[expected-pairs.csv](../metadata/expected-pairs.csv)
