---
output:
  html_document:
    toc: true
    toc_float: true
    highlight: zenburn
    theme: flatly
---

```{r read-data}
library(readr)
bc = read_delim("GEN00067700-barcodes.txt", delim=" ", progress=FALSE)
```

```{r barcode-group}
library(dplyr)
grouped = bc %>%
  group_by(barcode, index) %>%
  summarise(count=n(), bcq=mean(barcodeq), iq=mean(indexq))
```

# Summary stats
There are `r nrow(grouped)` unique barcode-index pairs and of those
`r nrow(subset(grouped, count > 10))` are seen more than 10 times.

We can see below that there is are a small set of barcode-index
pairs that are represented a high number of times.

```{r barcode-index-plot}
library(ggplot2)
ggplot(grouped, aes(count)) + geom_histogram() + theme_bw() +
  scale_x_log10() + scale_y_sqrt()
```

# Known barcodes
Now we'll mark up the barcode-index pair with whether or not these
are some of the known sequences. The indexes we found correspond to the
reverse complement column in the index file and the barcodes correspond
to the sense column in the barcode file.

```{r read-known-data}
kbc = read_csv("../metadata/barcode.csv")
kin = read_csv("../metadata/index.csv")
grouped$kin = grouped$index %in% kin$rc
grouped$kbc = grouped$barcode %in% kbc$anti_sense
grouped$known = grouped$kin & grouped$kbc
grouped$known = ifelse(grouped$known, "known", "unknown")
```

Here we do a bunch of transformations to get the expected index-barcode
pairs in a useful format. We take all the expected barcode matchings for
each experiment and each pool separately, and then determine which indexes
should have which barcodes. Then we combine them all together at the end.


```{r expected-barcode-index-pairs}
expected = read_csv("../metadata/GEN00067700-indexpairs.csv", progress=FALSE)
mcneu = read_tsv("../metadata/McNeu.tsv", progress=FALSE) %>%
  mutate(pool="McNeu Pool") %>%
  left_join(expected, by=c("pool"="bcid")) %>%
  select(-pool)
hmler = read_tsv("../metadata/HMLER.tsv") %>%
  mutate(pool="HMLER Pool") %>%
  left_join(expected, by=c("pool"="bcid")) %>%
  select(-pool)
poola = read_tsv("../metadata/PoolA.csv") %>%
  mutate(pool="Pool A") %>%
  left_join(expected, by=c("pool"="bcid")) %>%
  select(-pool)
expected = subset(expected, !bcid %in% c("McNeu Pool", "HMLER Pool", "Pool A"))
expected$bcid = as.integer(expected$bcid)
expected = expected %>% bind_rows(mcneu, hmler, poola) %>%
  left_join(kbc, by=c("bcid"="id")) %>%
  select(anti_sense.y, anti_sense.x)
colnames(expected) = c("ebarcode", "eindex")
expected$pair = paste(expected$ebarcode, expected$eindex, sep=":")
write_csv(expected, "expected-reformat.csv")
```

That was a bunch of joining and merging, so let's make sure we did it correctly.
We're expecting 10 indexes to have 2 barcodes each, 3 indexes to have 30
barcodes (HMLER Pool), 6 indexes to have 33 barcodes (McNeu pool) and 1 index to
have 6 barcodes (Pool A). That is a total of `r (10 * 2) + (3 * 30) + (6 * 33) +
(1 * 6)` barcode index pairs. There are `r nrow(expected)` rows in the expected
barcode data frame, once we are done with all the merging, so it looks like we
did the merging correctly.

```{r  barcodes-index-pairing}
grouped$pair = paste(grouped$barcode, grouped$index, sep=":")
grouped$expected = grouped$pair %in% expected$pair
known = subset(grouped, known == "known")
```

We only only see `r sum(grouped$expected)` of the `r nrow(expected)` expected
barcode-index pairs in this dataset. Some indexes are missing expected
barcodes:

```{r missing-barcodes-by-index}
missing_pairs = expected %>%
  mutate(present=pair %in% grouped$pair) %>%
  group_by(eindex, present) %>% summarise(count=n())
ggplot(missing_pairs, aes(eindex, count, color=present)) +
  geom_point() +
  xlab("") +
  theme(axis.text.x  = element_text(angle=90))
```

`ACGGTATA` looks like it is a failed library. There are only
`r sum(subset(grouped, index == "ACGGTATA")$count)` total reads matching
this index in the library.

I wrote out the barcode-index pairs with counts, the mean of the barcode quality (bcq),
the mean of the index quality (iq), whether or not the index is a known
index (kin), whether or not the barcode is a known barcode (kin), and
whether or not the barcode-index pair is expected (known).

This table is here [All barcode index](all-barcode-index-unfiltered.csv)

```{r barcode-out}
write.table(grouped, file="all-barcode-index-unfiltered.csv", col.names=TRUE, row.names=FALSE, quote=FALSE, sep=",")
```

Now we can see if we mark up the barcode-index pairs with whether or not
they are from the known barcode and known index files, we find that
there are `r nrow(known)` barcode-index pairs and `r nrow(subset(known, expected))`
expected barcode-index pairs.

If we plot the histograms separating out whether they are a known or unknown
barcode-index pair, we can see that the known barcode-index pairs
have many more reads associated with them than the unknown barcode-index
pairs.

```{r known-plot}
library(ggplot2)
ggplot(grouped, aes(count)) + geom_histogram() + theme_bw() +
  facet_wrap(~known) +
  scale_x_log10() + scale_y_sqrt() +
  ylab("number of barcode-index pairs")
```

We can see that the PHRED quality of the known barcode-index pairs has a tight
distribution:

```{r known-quality-plot}
ggplot(grouped, aes(bcq, iq)) + geom_point() + theme_bw() +
  facet_wrap(~known) +
  ylab("mean PHRED quality of index") +
  xlab("mean PHRED quality of barcode")
```

Here you can see that if we filter for only known barcodes and indexes, they
    tend to have both a high index and barcode quality compared to unknown
    sequences.

```{r known-bc-quality-plot}
ggplot(grouped, aes(bcq, iq, color=known)) + geom_point(size=0.5) + theme_bw() +
  facet_wrap(~kin+kbc) +
  ylab("mean PHRED quality of index") +
  xlab("mean PHRED quality of barcode")
```

If we look at the expected barcode-index pairings, we can see that there is not
much of a difference in the distribution of quality of the barcode-index pairs
for the pairs we expect (TRUE) and the pairs we do not expect (FALSE). Filtering
on quality will not help these results.

```{r expected-bc-quality}
ggplot(subset(grouped, known == "known"), aes(bcq, iq)) +
  geom_point(size=0.5) + theme_bw() +
  facet_wrap(~expected) +
  ylab("mean PHRED quality of index") +
  xlab("mean PHRED quality of barcode")
```

We can see that the barcode-index pairs have about an order of magnitude less
counts when they are not expected than when they are expected.

```{r expected-counts}
ggplot(subset(grouped, known == "known"), aes(count)) +
  geom_histogram() + theme_bw() +
  facet_wrap(~expected) +
  scale_x_log10()
```

# Barcodes per index
We generally see more barcodes than we are expecting per index.

```{r barcodes-per-index-plot}
expected_counts = expected %>% group_by(eindex) %>% summarise(nexpected=n())
barcode_per_index = known %>% group_by(index) %>% summarise(nbarcodes = n()) %>%
  left_join(expected_counts, by=c("index"="eindex")) %>%
  tidyr::gather("type", "count", 2:3)
index_per_barcode = known %>% group_by(barcode) %>% summarise(nindex = n())
ggplot(barcode_per_index, aes(index, count, fill=type)) +
  geom_bar(stat='identity', position='dodge') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  xlab("")
```

# Barcodes counts for each index

We see every index with every barcode at least once. Here we color
the plots by if they are the pair we expected or not.

```{r overrpresented}
ggplot(known, aes(barcode, count, fill=expected)) +
  facet_wrap(~index) + geom_bar(stat='identity') +
  scale_y_sqrt() +
  theme_bw() +
  theme(axis.ticks = element_blank(), axis.text.x = element_blank())
```

These aren't lining up at all with what we were expecting, it looks a lot worse
than the previous library. Let's pick one of the indexes that are supposed
to have just two barcodes "AATAGGAT" (top left) to look at more in depth.

```{r AATAGGAT-in-depth}
AATAGGAT = grouped %>%
  filter(index == "AATAGGAT")
AATAGGAT_known = subset(AATAGGAT, known == "known") %>% arrange(desc(count))
AATAGGAT_unknown = subset(AATAGGAT, known == "unknown")
```

There are `r nrow(AATAGGAT)` barcodes identified for AATAGGAT. Of those,
`r nrow(AATAGGAT_known)` have both a known barcode and a known index.
`r nrow(AATAGGAT_unknown)` do not have a known barcode and a known index.
For the known barcode-index pairs, there are `r sum(AATAGGAT_known$count)`
and for the uknown pairs there are `r sum(AATAGGAT_unknown$count)`, so despite
there being only a small number of entries where both the barcode and index
are known, they have the majority of the counts, which is what we expect.

Looking at the top of the table, we can see the two barcodes that are the
highest are not the barcodes we were expecting. We are expecting
barcodes 64 and 68 but the top two are barcodes 54 and 44.

```{r AATAGGAT-table, results='asis'}
knitr::kable(head(AATAGGAT_known))
```

I checked to see if we are making a mistake somewhere in this script by checking
in the unprocessed sequences by grepping for the barcode and the index in the
same read. This isn't perfect since it doesn't match the anchor sequences and
doesn't restrict the position the sequences are in, but gives a quick and dirty
estimate to make sure we didn't make a mistake anywhere:

```
TCATAATCTCAACAATCTTTCTTT (expected): 4006
CTTTTTCAATCACTTTCAATTCAT (top hit): 72315
```

# Output tables

The first table has the expected barcode-index pairings, reformatted to
be easy to use in the script.

[Expected barcode-index pairs file (reformatted)](expected-reformat.csv)

This table is all of the identified barcode-index pairs from the sequence
data (everything that matched the regex). It is unfiltered, and has columns
indicating whether the barcode was known 'kbc' the index was known 'kin',
both were known 'known' and if the barcode-index pair is expected 'expected'.
The count column is the total number of times we saw this barcode-index pairing.

[Barcode-index pairings (unfiltered)](all-barcode-index-unfiltered.csv)

# Summary
Something is up, we are seeing barcodes in all kinds of samples we are not
expecting. This isn't simple crosstalk, because if that was the case we would
see a low level of cross contamination but still see the main barcode signals
popping out like we did in the first set of samples. It is entirely possible
that I have screwed something up, but if I am, I can't figure out where it
is happening.
